Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-05-08 src/tex/ms.tex
change optimistic/PLATO sample size to "~" 300/500
to avoid confusion why we end up with 288 or 497.

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_statpwr_H2O-f.pdf}
        \caption{
            Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
            For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
            Higher values in either parameter result in a more reliable detection.
            For water mass fractions $\gtrsim 10^{-4}$, the statistical power largely depends on the fraction of greenhouse climate planets in the sample.
        }
        \label{fig:statpwr_H2O-f}
    \end{centering}
\end{figure}
Figure~\ref{fig:statpwr_H2O-f} explores the statistical power of the hypothesis test achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
It is highest for large water inventories and large dilution factors.
 For all but very low water fractions, $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
Within the framework of our model and as long as $f_\mathrm{rgh}$ is larger than $\sim 0.2$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $10^{-3}$.

@@

    \begin{figure*}[ht!]
%    \script{figurescript.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/plato_fgrid.pdf}
            \caption{
            Expected delta-evidences as a function of the fraction of planets with runaway greenhouse climates for different versions of the \plato\ survey.
            The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
            \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small dilution factors $\sim \var{platoMinfrgh}$ allow a detection.
                A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
                Without such follow-up measurements, sufficient diagnostic power can only be achieved with this sample if $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$.
               Even smaller samples are unlikely to yield a significant detection.
            \textit{Right:} Evidences when only FGK or only M~dwarfs are considered.
                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a reliable detection of the habitable zone inner edge discontinuity.
            }
            \label{fig:plato_fgrid}
        \end{centering}
    \end{figure*}
To simulate the transit survey of ESA's \plato\ mission, we considered a volume-limited sample with a size according to projections and including all stellar spectral types.
Given \plato's expected radius precision, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $ f_\mathrm{rgh} \sim \var{platoMinfrgh}$ (see Figure~\ref{fig:plato_fgrid}).
The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
For much smaller samples, only an optimistically strong signal is likely to be detected.

@@

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S_binned.pdf}
        \caption{
            Detection of the runaway greenhouse threshold with a binned statistic.
            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis. Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space. Random draws from the posterior of the runaway greenhouse hypothesis (Eqn.~\ref{eq:rgh_hypo}) are shown in red.
            As in the nominal case, the pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S_binned}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S_binned} shows the simulated data together with binned, average planet radii and draws from the posterior of the hypothesis test.
A clear detection resulted, although with somewhat lower significance ($\Delta \ln \mathcal{Z} \approx 30$) compared to the nominal setup.
The accuracy of the recovered instellation threshold is comparable.
This test demonstrates that our results are not sensitive on the choice of statistical estimator to test the hypotheses against.

>>>>

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N \approx \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N \approx \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_statpwr_H2O-f.pdf}
        \caption{
            Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
            For a sample size $N \approx \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
            Higher values in either parameter result in a more reliable detection.
            For water mass fractions $\gtrsim 10^{-4}$, the statistical power largely depends on the fraction of greenhouse climate planets in the sample.
        }
        \label{fig:statpwr_H2O-f}
    \end{centering}
\end{figure}
Figure~\ref{fig:statpwr_H2O-f} explores the statistical power of the hypothesis test achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
It is highest for large water inventories and large dilution factors.
 For all but very low water fractions, $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
Within the framework of our model and as long as $f_\mathrm{rgh}$ is larger than $\sim 0.2$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $10^{-3}$.

@@

    \begin{figure*}[ht!]
%    \script{figurescript.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/plato_fgrid.pdf}
            \caption{
            Expected delta-evidences as a function of the fraction of planets with runaway greenhouse climates for different versions of the \plato\ survey.
            The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
            \textit{Left:} For a large planet yield of $N \approx \var{N_plato}$, even small dilution factors $\sim \var{platoMinfrgh}$ allow a detection.
                A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
                Without such follow-up measurements, sufficient diagnostic power can only be achieved with this sample if $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$.
               Even smaller samples are unlikely to yield a significant detection.
            \textit{Right:} Evidences when only FGK or only M~dwarfs are considered.
                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a reliable detection of the habitable zone inner edge discontinuity.
            }
            \label{fig:plato_fgrid}
        \end{centering}
    \end{figure*}
To simulate the transit survey of ESA's \plato\ mission, we considered a volume-limited sample with a size according to projections and including all stellar spectral types.
Given \plato's expected radius precision, we find that a yield of $\sim$\var{N_plato} is needed for a significant detection if the fraction of runaway greenhouse planets $ f_\mathrm{rgh} = \var{platoMinfrgh}$ (see Figure~\ref{fig:plato_fgrid}).
The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
For much smaller samples, only an optimistically strong signal is likely to be detected.

@@

\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S_binned.pdf}
        \caption{
            Detection of the runaway greenhouse threshold with a binned statistic.
            Using simulated data from the optimistic survey case ($N \approx \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis. Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space. Random draws from the posterior of the runaway greenhouse hypothesis (Eqn.~\ref{eq:rgh_hypo}) are shown in red.
            As in the nominal case, the pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S_binned}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S_binned} shows the simulated data together with binned, average planet radii and draws from the posterior of the hypothesis test.
A clear detection resulted, although with somewhat lower significance ($\Delta \ln \mathcal{Z} \approx 30$) compared to the nominal setup.
The accuracy of the recovered instellation threshold is comparable.
This test demonstrates that our results are not sensitive on the choice of statistical estimator to test the hypotheses against.
